---
title: "Medical Appointment Scheduling System ver.2_datacleaning"
author: "Jaehui Park"
date: "2026-02-24"
output: html_document
---

# Data Source: Kaggle (Carolina Gonzalez Galtier)
# License: CC BY-NC-SA 4.0
# Description: Analysis of patient no-show patterns and seasonal trends.

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE,
  message = FALSE, 
  warning = FALSE
)
```

Library
```{r Library}
library(tidyverse)

library(readr)

#install.packages("psych")
library(psych)

library(dplyr)

library(lubridate)

library(RColorBrewer)

library(scales)

```


1. Data 

Data Loading
```{r Data Loading}

# Load data using relative paths for portability
# This ensures the code runs on any machine as long as the data folder exists
slots <- read_csv("data/slots.csv")
patients <- read_csv("data/patients.csv")
appointments <- read_csv("data/appointments.csv")
```

Data quality check: Primary Key Uniqueness
```{r Data Quality Check}
## Data Quality Check: Primary Key Uniqueness
nrow(appointments) == n_distinct(appointments$appointment_id)
nrow(slots) == n_distinct(slots$slot_id)
nrow(patients) == n_distinct(patients$patient_id)

```

Foreign key integrity
```{r Foreign Key Integrity}
## Data Quality Check: Foreign Key Integrity A.patient_ID = P.patient_ID

orphan_appointments <- anti_join(
  x = appointments,
  y = patients,
  by = "patient_id"
)

nrow(orphan_appointments)

```


```{r Foreign Key Integrity2}
## Data Quality Check: Foreign Key Integrity A.slot_ID = S.slot_ID

orphan_slots <- anti_join(
  x = appointments,
  y = slots,
  by = "slot_id"
)

nrow(orphan_slots)

```

```{r Missing Values check}
## Data Quality Check: Missing Values

check_na <- function(df, table_name) {
  cat("\n--- [", table_name, "] Missing Values ---\n")
  colSums(is.na(df)) %>% print()
}

# Apply to core tables
check_na(slots, "Slots")
check_na(patients, "Patients")
check_na(appointments, "Appointments")

```


```{r Visualize Appointment Status Counts}

appointments %>%
  count(status) %>%
  ggplot(aes(x = reorder(status, n), y = n, fill = status)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 4) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Appointment Status Distribution", x = "Status", y = "Count") +
  theme_minimal()

```

Total Missing: 25,456
Total (cancelled + did not attend + unknown + scheduled): 25,456
Conclusion: Data is consistent and clean.


The distribution of appointment statuses highlights the number of cancelled and no-show appointments, which can inform follow-up strategies and operational planning.

```{r Master dataset, full_dataset}
#Create a master dataset by joining appointment, patient, and slot data
full_dataset <- appointments %>%
  # Join patient-level information using patient_id
  left_join(patients, by = "patient_id", suffix = c("", "_pat")) %>%
  # Join slot-level information using slot_id
  left_join(slots, by = "slot_id", suffix = c("", "_slot"))

# Preview the structure of the combined dataset
glimpse(full_dataset)

```


Check full_dataset
```{r Check full_dataset}
# Validate that the row count remains consistent after joining
# The result should be TRUE
nrow(appointments) == nrow(full_dataset)

# Display row counts before and after the join for verification
cat("Number of appointments:", nrow(appointments), "\n")
cat("Number of rows of full_dataset:", nrow(full_dataset), "\n")

```

```{r check full_dataset2}
# Check for missing values introduced after the joins
full_dataset %>%
  summarise(
    missing_names = sum(is.na(name)),        # Unmatched patient records
    missing_slots = sum(is.na(is_available)) # Unmatched slot records
  )

```


```{r check full_dataset3}
# Create a validation summary to review the join results
join_validation <- data.frame(
  Table = c("Appointments (Original)", "full_dataset(After join)"),
  Row_Count = c(nrow(appointments), nrow(full_dataset)),
  Col_Count = c(ncol(appointments), ncol(full_dataset))
)

print(join_validation)

```


```{r}
# Identify new columns added to the dataset after the joins
setdiff(colnames(full_dataset), colnames(appointments))

```
Data Merging Summary: Row count consistency (111,488) and column expansion (16 → 23) adding the following 7 new columns to the master dataset.

```{r}
# List all column names in the final merged dataset
colnames(full_dataset)
```

Validated column redundancy: Merged fields with 100% identical values were removed to eliminate data duplication and optimize the master dataset

```{r}
# Check if redundant columns are 100% identical
identical(full_dataset$sex, full_dataset$sex_pat)
identical(full_dataset$appointment_date, full_dataset$appointment_date_slot)
identical(full_dataset$appointment_time, full_dataset$appointment_time_slot)
```
Confirmed 100% matches for redundant fields.
Removed sex_pat, appointment_date_slot, and appointment_time_slot to optimize the master dataset.

```{r}
# Remove redundant columns after identity verification
full_dataset <- full_dataset %>%
  select(-sex_pat, -appointment_date_slot, -appointment_time_slot)

# Final check of the streamlined column list
colnames(full_dataset)
```



```{r}
# 1. Set the reference date (December 1, 2024)
ref_date <- as.Date("2024-12-01")

# 2. Past appointments (before the reference date)
# Used for analyzing actual visits, cancellations, and no-shows
past_appointments <- full_dataset %>%
  filter(appointment_date < ref_date)

# 3. Future appointments (on or after the reference date)
# Used for forecasting hospital utilization or appointment fulfillment
future_appointments <- full_dataset %>%
  filter(appointment_date >= ref_date)

# 4. Verify the data split
cat("Total rows in full dataset:", nrow(full_dataset), "\n")
cat("Rows in past appointments (analysis):", nrow(past_appointments), "\n")
cat("Rows in future appointments (forecasting):", nrow(future_appointments), "\n")

```


The dataset was split into 'Past' and 'Future' records based on the reference date (December 1, 2024). Past data is used for analyzing historical performance, while future data is reserved for forecasting hospital demand.


Check unique values for each relevant column

```{r}
# Identify and display unique values for all categorical columns at once
# We select columns of type 'character' or 'factor' and map the unique function
past_appointments %>%
  select(status, sex, age_group, insurance) %>% # Select relevant categorical columns
  map(unique) # Apply the unique() function to each selected column
```
Inspected unique values for categorical columns to understand the data distribution and identify distinct classes.


```{r}
# 1. Define the logical order for age-based categories
# This prevents alphabetical sorting and maintains the natural age sequence
age_levels <- c("15-19", "20-24", "25-29", "30-34", "35-39", "40-44", 
                "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", 
                "75-79", "80-84", "85-89", "90+")

# 2. Transform character columns into structured factors
# Assigning specific levels ensures consistent categorization and plotting order
past_appointments <- past_appointments %>%
  mutate(
    status = factor(status, levels = c("attended", "did not attend", "cancelled", "unknown")),
    sex = factor(sex, levels = c("Male", "Female")),
    age_group = factor(age_group, levels = age_levels),
    insurance = as.factor(insurance)
  )

# 3. Validate the distribution of categorical variables
# Summary() now provides frequency counts for each factor level
summary(past_appointments %>% select(status, sex, age_group, insurance))

```


Saving cleaned data, past_appointments
```{r}
# Create a dedicated directory for data storage if it doesn't already exist
if (!dir.exists("data")) {
  dir.create("data")
}

# Export the refined 'past_appointments' dataset to an RDS file
# This ensures that all categorical sorting and data structures remain intact
saveRDS(past_appointments, "data/cleaned_past_appointments.rds")
```





